---
title: "ST558 Project 1"
author: "Maggie Feng"
date: "June 8, 2021"
output: 
  github_document:
    toc: true
---

Note: required packages to run the code and create vignette

```{r}
knitr::render_markdown()
```

# National Hockey League (NHL) Vignette

## NHL records API

The `getRecords` function will contact the NHL records API and pull data from one of the six enpoint and output the data as a data frame.

The first argument `input` is to specify one of the six enpoints: 

*`"franchise"` will return id, firstSeasonId, lastSeasonId, and name of every team in the history of the NHL
*`"teamtotal"` will return the total stats for every franchise
*`"season"` will drill-down into season records and can do so for a specific franchise
*`"goalie"` will return goalie records for all or a specified franchise
*`"skater"` will return skater records for all or a specified franchise
*`"history"` will return admin history and retired numbers

The arguments `name` and `id` are used to specify which team or franchise's records you wish to output.

```{r}
#Required packages
library(httr)
library(jsonlite)

#Function to contact the NHL records API
#returns well-formatted, parsed data
getRecords <- function(input = c("franchise", "teamtotal", "season", "goalie", "skater", "history"), 
                       name = NULL, id = NULL){
  ##Note: if both team name and id arguments are filled, team name overrides id number
  
  #All queries are prefixed by this base_url
  base_url <- "https://records.nhl.com/site/api"
  
  #First two endpoints: do not have name/id specification
  if (input == "franchise" | input == "teamtotaol"){
    if (input == "franchise")
      endpoint <- "franchise"
    if (input == "teamtotal")
      endpoint <- "franchise-team-totals"
    full_url <- paste0(base_url, "/", endpoint)
  }
  
  #Third, fourth, and fifth endpoints: with name/id specification by franchiseId
  if (input == "season" | input == "goalie" | input == "skater"){
    if (input == "season")
      endpoint <- "franchise-season-records"
    if (input == "goalie")
      endpoint <- "franchise-goalie-records"
    if (input == "skater")
      endpoint <- "franchise-skater-records"
    
    if (is.null(name) & is.null(id)){
      full_url <- paste0(base_url,"/", endpoint)
    }
    else {
      if (!is.null(id)) {
        full_url <- paste0(base_url,"/", endpoint, "?cayenneExp=franchiseId=", id)
      }
      if (!is.null(name)) {
        v <- c("Montreal Canadiens", "Montreal Wanderers", "St. Louis Eagles", "Hamilton Tigers",
               "Toronto Maple Leafs", "Boston Bruins", "Montreal Maroons", "Brooklyn Americans",
               "Philadelphia Quakers", "New York Rangers", "Chicago Blackhawks", "Detroit Red Wings",
               "Cleveland Barons", "Los Angeles Kings", "Dallas Stars", "Philadelphia Flyers", 
               "Pittsburgh Penguins", "St. Louis Blues", "Buffalo Sabres", "Vancouver Canucks", 
               "Calgary Flames", "New York Islanders", "New Jersey Devils", "Washington Capitals",
               "Edmonton Oilers", "Carolina Hurricanes", "Colorado Avalanche", "Arizona Coyotes",
               "San Jose Sharks", "Ottawa Senators", "Tampa Bay Lightning", "Anaheim Ducks",
               "Florida Panthers", "Nashville Predators", "Winnipeg Jets", "Columbus Blue Jackets",
               "Minnesota Wild", "Vegas Golden Knights", "Seattle Kraken")
        nameid <- match(name, v)
        full_url <- paste0(base_url,"/", endpoint,"?cayenneExp=franchiseId=", nameid)
      }
    }
  }
  
  #Sixth endpoint: has name/id specification by mostRecentTeamId instead of franchiseId
  if (input == "history"){
    endpoint <- "franchise-detail"
    if (is.null(name) & is.null(id)){
      full_url <- paste0(base_url, "/", endpoint)
    }
    else {
      if (!is.null(id)) {
        full_url <- paste0(base_url, "/", endpoint, "?cayenneExp=mostRecentTeamId=", id)
      }
      if (!is.null(name)) {
        u <- c("New Jersey Devils", "New York Islanders", "New York Rangers", "Philadelphia Flyers",
               "Pittsburgh Penguins", "Boston Bruins", "Buffalo Sabres", "Montreal Canadiens",
               "Ottawa Senators", "Toronto Maple Leafs", "Carolina Hurricanes", "Florida Panthers",
               "Tampa Bay Lightning", "Washington Capitals", "Chicago Blackhawks", 
               "Detroit Red Wings", "Nashville Predators", "St. Louis Blues", "Calgary Flames", 
               "Colorado Avalanche", "Edmonton Oilers", 
               "Vancouver Canucks", "Anaheim Ducks", "Dallas Stars",
               "Los Angeles Kings", "San Jose Sharks", "Columbus Blue Jackets", "Minnesota Wild",
               "Hamilton Tigers", "Philadelphia Quakers", "Montreal Wanderers", "Montreal Maroons",
               "St. Louis Eagles", "Cleveland Barons", "Brooklyn Americans", "Winnipeg Jets",
               "Arizona Coyotes", "Vegas Golden Knights", "Seattle Kraken")
        nameid <- match(name, u)
        unum <- c(1:10, 12:30, seq(37,45,by=2), 49, 51:55)
        full_url <- paste0(base_url,"/", endpoint, "?cayenneExp=mostRecentTeamId=", unum[nameid])
      }
    }
  }
  x <- GET(full_url)           #Retrieving data from URL
  y <- content(x, as = "text") #Converiting it to JSON text
  z <- fromJSON(y)             #Converting to list
  as.data.frame(z$data)        #Converting to data frame
}

## Examples:
#Francise information
franchise <- getRecords("franchise")

#Season records for specific franchise
season <- getRecords("season", id=1)

#Goalie records for specific franchise
goalie <- getRecords("goalie", name="Carolina Hurricanes")

#Admin history and retired numbers, Note: id is mostRecentTeamId
admin <- getRecords("history", id=1)
```

## NHL stats API

The `getStats` function

```{r}
getStats <- function(input, id = NULL){
  
}
```

