---
title: "National Hockey League (NHL) Vignette"
author: "Maggie Feng"
date: "June 20, 2021"
output: 
  github_document:
    toc: true
---

```{r setup, include = FALSE}
#README file
#knitr::render(input="ST558Project1.Rmd", output = "readme.md")
#knit(input="ST558Project1.Rmd", output = "readme.md")

#Note: required packages to run the code and create vignette
#devtools::build_vignettes()
```

Required packages to run this vignette

```{r, message = FALSE}
#Required packages
library(httr)
library(jsonlite)
library(tidyverse)
```

# Functions to read in the data

## NHL records API

The `getRecords` function will contact the [NHL records API](https://gitlab.com/dword4/nhlapi/-/blob/master/records-api.md), pull data from one of the six enpoints, and output the data as a data frame.

The first argument `input` is to specify one of the six endpoints: 

* `"franchise"` will return id, firstSeasonId, lastSeasonId, and name of every team in the history of the NHL
* `"teamtotal"` will return the total stats for every franchise
* `"season"` will drill-down into season records and can do so for a specific franchise
* `"goalie"` will return goalie records for all or a specified franchise
* `"skater"` will return skater records for all or a specified franchise
* `"history"` will return admin history and retired numbers

The arguments `name` and `id` are used to specify which team or franchise's records you wish to output. Adding `name` or `id` for the first two endpoints, `"franchise"` and `"teamtotal"`, will have the same result of not adding the arguments at all. 

```{r}
#Function to contact the NHL records API
#returns well-formatted, parsed data
getRecords <- function(input = c("franchise", "teamtotal", "season", "goalie", "skater", "history"), 
                       name = NULL, id = NULL){
  ##Note: if both team name and id arguments are filled, team name overrides id number
  
  #All queries are prefixed by this base_url
  base_url <- "https://records.nhl.com/site/api"
  
  #First two endpoints: do not have name/id specification
  if (input == "franchise" | input == "teamtotal"){
    if (input == "franchise")
      endpoint <- "franchise"
    if (input == "teamtotal")
      endpoint <- "franchise-team-totals"
    full_url <- paste0(base_url, "/", endpoint)
  }
  
  #Third, fourth, and fifth endpoints: with name/id specification by franchiseId
  if (input == "season" | input == "goalie" | input == "skater"){
    if (input == "season")
      endpoint <- "franchise-season-records"
    if (input == "goalie")
      endpoint <- "franchise-goalie-records"
    if (input == "skater")
      endpoint <- "franchise-skater-records"
    
    if (is.null(name) & is.null(id)){
      full_url <- paste0(base_url,"/", endpoint)
    }
    else {
      if (!is.null(id)) {
        full_url <- paste0(base_url,"/", endpoint, "?cayenneExp=franchiseId=", id)
      }
      if (!is.null(name)) {
        v <- c("Montreal Canadiens", "Montreal Wanderers", "St. Louis Eagles", "Hamilton Tigers",
               "Toronto Maple Leafs", "Boston Bruins", "Montreal Maroons", "Brooklyn Americans",
               "Philadelphia Quakers", "New York Rangers", "Chicago Blackhawks", "Detroit Red Wings",
               "Cleveland Barons", "Los Angeles Kings", "Dallas Stars", "Philadelphia Flyers", 
               "Pittsburgh Penguins", "St. Louis Blues", "Buffalo Sabres", "Vancouver Canucks", 
               "Calgary Flames", "New York Islanders", "New Jersey Devils", "Washington Capitals",
               "Edmonton Oilers", "Carolina Hurricanes", "Colorado Avalanche", "Arizona Coyotes",
               "San Jose Sharks", "Ottawa Senators", "Tampa Bay Lightning", "Anaheim Ducks",
               "Florida Panthers", "Nashville Predators", "Winnipeg Jets", "Columbus Blue Jackets",
               "Minnesota Wild", "Vegas Golden Knights", "Seattle Kraken")
        nameid <- match(name, v)
        full_url <- paste0(base_url,"/", endpoint,"?cayenneExp=franchiseId=", nameid)
      }
    }
  }
  
  #Sixth endpoint: has name/id specification by mostRecentTeamId instead of franchiseId
  if (input == "history"){
    endpoint <- "franchise-detail"
    if (is.null(name) & is.null(id)){
      full_url <- paste0(base_url, "/", endpoint)
    }
    else {
      if (!is.null(id)) {
        full_url <- paste0(base_url, "/", endpoint, "?cayenneExp=mostRecentTeamId=", id)
      }
      if (!is.null(name)) {
        u <- c("New Jersey Devils", "New York Islanders", "New York Rangers", "Philadelphia Flyers",
               "Pittsburgh Penguins", "Boston Bruins", "Buffalo Sabres", "Montreal Canadiens",
               "Ottawa Senators", "Toronto Maple Leafs", "Carolina Hurricanes", "Florida Panthers",
               "Tampa Bay Lightning", "Washington Capitals", "Chicago Blackhawks", 
               "Detroit Red Wings", "Nashville Predators", "St. Louis Blues", "Calgary Flames", 
               "Colorado Avalanche", "Edmonton Oilers", "Vancouver Canucks", "Anaheim Ducks", 
               "Dallas Stars","Los Angeles Kings", "San Jose Sharks", "Columbus Blue Jackets", 
               "Minnesota Wild", "Hamilton Tigers", "Philadelphia Quakers", "Montreal Wanderers",
               "Montreal Maroons","St. Louis Eagles", "Cleveland Barons", "Brooklyn Americans",
               "Winnipeg Jets", "Arizona Coyotes", "Vegas Golden Knights", "Seattle Kraken")
        nameid <- match(name, u)
        unum <- c(1:10, 12:30, seq(37,45,by=2), 49, 51:55)
        full_url <- paste0(base_url,"/", endpoint, "?cayenneExp=mostRecentTeamId=", unum[nameid])
      }
    }
  }
  x <- GET(full_url)           #Retrieving data from URL
  y <- content(x, as = "text") #Converiting it to JSON text
  z <- fromJSON(y)             #Converting to list
  as.data.frame(z$data)        #Converting to data frame
}
```

Below are some examples of how to use the function. The first outputs all of the franchise information. The second shows how adding an `id` (or `name`) argument will not affect the `"teamtotal"` (and `"franchise"`) inputs. The third shows all season records, but can also have a specification of `id` or `name`. The fourth shows goalie information for the team with 1 as the `franchiseId`. The fifth shows the skater inforamtion for the team with `name` Carolina Hurricanes. The sixth shows the admin history and retired numbers for the team with 1 as the `mostRecentTeamId`.

```{r, message = FALSE}
## Examples:
#Franchise information
franchise <- getRecords("franchise")

#Total stats for every franchise
total <- getRecords("teamtotal", id=1)

#Season records for specific franchise or for all 
season <- getRecords("season")

#Goalie records for specific franchise
goalie <- getRecords("goalie", id=1)

#Skater records for for specific franchise
skater <- getRecords("skater", name = "Carolina Hurricanes")

#Admin history and retired numbers, Note: id is mostRecentTeamId
admin <- getRecords("history", id=1)
```

## NHL stats API

The `getStats` function will contact the [NHL stats API](https://gitlab.com/dword4/nhlapi/-/blob/master/stats-api.md) for the modifier and will output a data frame for a specified team or all teams. The arguments `recentId` and `name` can be specified if we want to look at a specific team. Otherwise, we can leave the arguments empty to output all teams stats.

```{r}
getStats <- function(name = NULL, recentId = NULL){
  base_url <- "https://statsapi.web.nhl.com/api/v1/teams"
  
  if (is.null(name) & is.null(recentId)){
      full_url <- paste0(base_url, "?expand=team.stats")
      x <- GET(full_url)
      y <- content(x, as = "text")
      z <- fromJSON(y)
      df <- NULL
      for (i in 1:31){
        a <- z$teams[i,]$teamStats[[1]]$splits[[1]]$stat[1,]
        b <- z$teams[i,][,c("id","name")]
        c <- cbind(b,a)
        df <- rbind(df,c)
      }
      df
    }
    else {
      if (!is.null(recentId)) {
        full_url <- paste0(base_url, "/", recentId, "?expand=team.stats")
      }
      if (!is.null(name)) {
        u <- c("New Jersey Devils", "New York Islanders", "New York Rangers", "Philadelphia Flyers",
               "Pittsburgh Penguins", "Boston Bruins", "Buffalo Sabres", "Montreal Canadiens",
               "Ottawa Senators", "Toronto Maple Leafs", "Carolina Hurricanes", "Florida Panthers",
               "Tampa Bay Lightning", "Washington Capitals", "Chicago Blackhawks", 
               "Detroit Red Wings", "Nashville Predators", "St. Louis Blues", "Calgary Flames", 
               "Colorado Avalanche", "Edmonton Oilers", "Vancouver Canucks", "Anaheim Ducks", 
               "Dallas Stars","Los Angeles Kings", "San Jose Sharks", "Columbus Blue Jackets", 
               "Minnesota Wild", "Hamilton Tigers", "Philadelphia Quakers", "Montreal Wanderers",
               "Montreal Maroons","St. Louis Eagles", "Cleveland Barons", "Brooklyn Americans",
               "Winnipeg Jets", "Arizona Coyotes", "Vegas Golden Knights", "Seattle Kraken")
        nameid <- match(name, u)
        unum <- c(1:10, 12:30, seq(37,45,by=2), 49, 51:55)
        full_url <- paste0(base_url, "/", unum[nameid], "?expand=team.stats")
      }
      x <- GET(full_url)
      y <- content(x, as = "text")
      z <- fromJSON(y)
      a <- z$teams[1,]$teamStats[[1]]$splits[[1]]$stat[1,]
      b <- z$teams[1,][,c("id","name")]
      cbind(b,a)
    }
}
```

Below are examples of calling the `getStats` function. The first calls all teams and the second calls the first team.

```{r}
## Examples
#All stats for all teams
all <- getStats()

#For a specific team
team1 <- getStats(recentId = 1)
team2 <- getStats(name = "Dallas Stars")
```

## API wrapper

The `getEndpoint` function is a wrapper function that allows the user to access any of the six API endpoints, including modifiers, teamIDs, etc. The function has arguments: 
* `input` to specify which of the six endpoints is to be used
* `name` because of the differences in `id` and `mostRecentId` only the name should be specified as the teamID
* `modifier` is set to `FALSE` but can be set to `TRUE` to obtain the NHL stats API modifier

```{r}
getEndpoint <- function(input = NULL, name, modifier = FALSE){
  getRecords(input, name = NULL, id = NULL)
  if (modifier = TRUE){
    getStats(name = NULL)
  }
}
```

# Exploratory Data Analysis

## Bar plot

## Histogram

## Box plot

## Scatter plot
